The Unicode Cookbook \linebreak for Linguists

### 2018-01-13 book/chapters/ipa_background.tex

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single ``correct'' transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a 
`narrow' phonetic transcription and a `broad' phonological transcription. 
A `narrow' phonetic transcription may freely use the all symbols in the IPA 
chart with direct reference to their phonetic value, i.e. the transcriber can 
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
is a simultaneous labial and velar closure which is voiced and contains nasal 
airflow, independently of the phonemic status of this sound in the language in 
question.

@@

In contrast, the basic goal of a phonemic ``broad'' transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).} 
\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized ``computer
codes'' to phonetic symbols was assigned to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives <~ƥ,~ƭ,~ƈ,~ƙ,~ʠ~> were added; in the 1993 
revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA Number and 
an IPA Name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
the \textsc{IPA extensions} block.

The IPA Number is simply expressed as a ``three-digit number numerical directory
of digit triples'' \cite{}. The numbering scheme specifies three-digit codes,
the first digit of which indicates the symbol's category
\citep{Esling1990,EslingGaylord1993}:

@@

In the historical context of the IPA being considered for inclusion in 
ISO/IEC 10646, it is important to realize that there were a variety of 
sources (i.e.\ not just from the Association) which submitted character 
proposals for phonetic alphabets. These proposals, including the one from the 
Association via the Kiel working group, were considered as a whole by 
the ISO working groups that were responsible for incorporating a phonetic 
script into the universal character set (UCS). The ISO working groups that 
were responsible for assigning a phonetic character set then made their 
own submissions as part of a review process by ISO for approval based on 
both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.

>>>>

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single correct transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a 
`narrow' phonetic transcription and a `broad' phonological transcription. 
A `narrow' phonetic transcription may freely use the all symbols in the IPA 
chart with direct reference to their phonetic value, i.e. the transcriber can 
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
is a simultaneous labial and velar closure which is voiced and contains nasal 
airflow, independently of the phonemic status of this sound in the language in 
question.

@@

In contrast, the basic goal of a phonemic broad transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).} 
\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was assigned to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives <~ƥ,~ƭ,~ƈ,~ƙ,~ʠ~> were added; in the 1993 
revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
recognized IPA symbols; they are assigned to the 200 series of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA Number and 
an IPA Name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
the \textsc{IPA extensions} block.

The IPA Number is simply expressed as a three-digit number numerical directory
of digit triples. The numbering scheme specifies three-digit codes,
the first digit of which indicates the symbol's category
\citep{Esling1990,EslingGaylord1993}:

@@

In the historical context of the IPA being considered for inclusion in 
ISO/IEC 10646, it is important to realize that there were a variety of 
sources (i.e.\ not just from the Association) which submitted character 
proposals for phonetic alphabets. These proposals, including the one from the 
Association via the Kiel working group, were considered as a whole by 
the ISO working groups that were responsible for incorporating a phonetic 
script into the universal character set (UCS). The ISO working groups that 
were responsible for assigning a phonetic character set then made their 
own submissions as part of a review process by ISO for approval based on 
both informatic and phonetic criteria \citep[86]{EslingGaylord1993}.

### 2018-01-13 book/chapters/ipa_meets_unicode.tex

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode Standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

>>>>

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.
